\textbf{Reference-based OTU clustering for machine learning classification}

### 2022-06-20 submission/manuscript.tex
revised ms

Other Title Options:\\
A solution to the problem of inconsistent OTU clustering for machine
learning classification

Optimized reference-based OTU clustering for machine learning
classification

@@

\hypertarget{abstract-250-word-max}{%
\subsection{Abstract (250 word max)}\label{abstract-250-word-max}}

Machine learning classification of disease based on the gut microbiome
often relies on operational taxonomic units (OTUs) to quantify microbial
composition. The standard methodology for quantifying taxonomic
composition is clustering 16S rRNA amplicon sequences \emph{de novo}
into OTUs and using their abundances to train a classification model.
However, OTU clusters depend on the sequences in the data set and
therefore can change if new data is added. This inconsistency
complicates machine learning classification because in order to use the
model to classify additional samples, the OTUs have to be re-clustered
to include the new sequences and the model retrained with the new OTU
clusters. A new algorithm in mothur, called OptiFit, addresses this
issue by clustering new sequences into the existing OTU clusters. While
OptiFit is proven to produce high quality OTU clusters, it is unclear
whether this method for clustering new sequence data into existing OTUs
will have any impact on machine learning classification using the OTU
abundances. To evaluate the potential for use in machine learning
classification, we used OptiFit to cluster additional data into existing
OTU clusters and quantified model performance in classifying a data set
containing samples from controls and people with advanced adenoma and
carcinoma (SRN). We compared the model performance using OptiFit to the
standard procedure of clustering all the data together. We found that
OptiFit can be used to fit new sequence data to existing OTUs and
perform equally in classifying SRNs. Moving forward, when OTUs are used
in machine learning classification problems, OptiFit can be used to
avoid the need to re-cluster sequences into OTUs when classifying
additional samples that were not part of the data set used to build the
model.

\hypertarget{importance-150-word-max}{%
\subsection{Importance (150 word max)}\label{importance-150-word-max}}

There are instances when OTU abundance data is optimal for machine
learning prediction of disease based on microbiome composition. The
current methodology for OTU clustering complicates machine learning
since \emph{de novo} OTU clusters can change depending on the data in
the data set. Attempting to classify additional samples that were not
part of the original model build requires re-clustering OTUs and
re-training the model. OptiFit provides an elegant solution to the
problem of OTU inconsistency by enabling new data to be fit to the
existing OTUs while maintaining machine learning classification
performance. The use of OptiFit will enable a single model based on OTU
abundance data to be used in classifying additional samples that were
not part of the original OTU clustering.

@@

Gut microbiome community composition is useful as a resource for machine
learning prediction of diseases, including colorectal cancer
(\protect\hyperlink{ref-baxter2016}{1},
\protect\hyperlink{ref-zackular2014}{2}). Amplicon sequencing of the 16S
rRNA gene is a reliable tool for assessing the taxonomic composition of
microbial communities. Analysis of 16S rRNA amplicon sequence (16S) data
generally relies on clustering of sequences based on similarity into
operational taxonomic units (OTUs). The process of OTU clustering can
either be reference-based or \emph{de novo}. The quality of OTUs
generated with reference-based clustering is dependent on the quality of
the reference used, while OTUs generated with \emph{de novo} clustering
are optimized to the data set (\protect\hyperlink{ref-westcott2015}{3}).
While \emph{de novo} clustering can produce high-quality OTU clusters
where sequences are accurately grouped based on similarity thresholds,
the resulting OTU clusters depend on the data in the data set and the
addition of new data could change the overall OTU clusters. The
inconsistent nature of OTU clustering complicates deployment of machine
learning models since integration of additional data requires
re-clustering all the data and re-training of the model. The ability to
integrate new data into an existing model without re-clustering and
re-training could allow for deployment of a single model that new data
can be continually added to. Recently Sovacool \emph{et al} introduced
OptiFit: a method for fitting new data into existing OTU clusters
(\protect\hyperlink{ref-sovacool2022}{4}). While OptiFit is proven to
effectively fit new sequence data to existing OTU clusters, it is
unknown if the use of OptiFit will have an impact on machine learning
classification. Here we test the ability of OptiFit to cluster new
sequence data into existing OTU clusters for the purpose of machine
learning classification of disease based on gut microbiome composition.

When using OTU abundances for machine learning classification, the
current methodology is to \emph{de novo} cluster all of the sequence
data into OTUs with the OptiClust algorithm in mothur
(\protect\hyperlink{ref-westcott2017}{5}). The resulting abundance data
is then split into training and testing sets, where the training set is
used to tune hyperparameters and ultimately train the model. The testing
set is then classified with the model and the performance of the model
can be quantified (Figure 1A). However, with this methodology we would
have to re-generate the OTU clusters and re-train the model if we wanted
to classify additional samples. The OptiFit algorithm
(\protect\hyperlink{ref-sovacool2022}{4}) addresses this problem by
enabling new sequences to be clustered into existing OTUs. The OptiFit
workflow is similar to the OptiClust workflow where the data is
clustered into OTUs and used to tune hyperparameters and ultimately
train the model. Then, we can use OptiFit to fit sequence data of
samples not part of the original data set into the existing OTUs and use
the same model to classify the samples (Figure 1B). To test how the
model performance compares between these two methodologies, we used a
publicly available data set of 16S sequences from stool samples of
healthy subjects as well as subjects with SRN consisting of advanced
adenoma and carcinoma (\protect\hyperlink{ref-baxter2016}{1}). The data
set was randomly split into an 80\% train set and 20\% test set. For the
standard OptiClust workflow, the training and test sets were \emph{de
novo} clustered together into OTUs then the resulting abundance table
was split into the training and testing set. For the OptiFit workflow,
the train set was clustered \emph{de novo} into OTUs and the remaining
test set was fit to the OTU clusters using the OptiFit algorithm. For
both workflows, the abundance table of the train set was used to tune
hyperparameters and train a random forest model to classify SRN. The
test set was classified as either control or SRN using the trained
models.To account for variation depending on the split of the data, the
data set was randomly split 100 times and the process repeated for each
of the 100 data splits. By comparing the model performance of
classifying the samples in the test data set between the OptiFit and
OptiClust algorithms, we can quantify the impact of using OptiFit on
model classification performance.

We first examined the quality of the resulting OTU clusters from the two
algorithms using the Matthews correlation coefficient (MCC). The MCC
score is quantified by examining all pairs of sequences and assessing
whether they belong together in an OTU or not based on sequence
similarity (\protect\hyperlink{ref-westcott2017}{5}). MCC scores range
between negative one and one. A score of negative one means none of the
sequences in an OTU are within the similarity threshold and any
sequences within the similarity threshold are not in an OTU together. An
MCC score of zero essentially means the sequences are randomly
clustered. An MCC score of 1 means all sequences in an OTU are within
the similarity threshold and all sequence pairs within the similarity
threshold are in the same OTU. To ensure that OptiFit is appropriately
integrating new sequence data into the existing OTUs, we would expect
that the MCC scores produced by the OptiClust and OptiFit workflows are
similar. Since the data is only clustered once in the OptiClust workflow
there is only one MCC score while the OptiFit workflow produces an MCC
score for the OTU clusters from each data split. Overall the MCC scores
were similar between OptiClust (MCC = 0.884) and OptiFit (average MCC =
0.879) indicating that OptiFit performs as well as OptiClust when
integrating new sequences into the existing OTUs.

After verifying that the quality of the OTUs was consistent between
OptiClust and OptiFit, we next examined the model performance for
classifying samples in the test data set as control or SRN. To quantify
model performance we used the taxonomic abundances of the training data
from the OptiClust and OptiFit workflows to train a model to predict
SRNs. Using the predicted and actual diagnosis classification, we
calculated the area under the receiver operating characteristic curve
(AUROC) for each data split to quantify model performance. During
cross-validation (CV) training, the model performance was equivalent
between the two algorithms (p-value = 0.13, OptiClust mean CV AUROC =
0.694, OptiFit mean CV AUROC = 0.697, Figure 2A). The trained model was
then deployed to classify the samples of the test data as control or
SRN. The performance on the test data was equivalent between the two
algorithms (p-value = 0.63, OptiClust mean test AUROC = 0.709, OptiFit
mean test AUROC = 0.712, Figure 2B,C) indicating that new data can be
fit to existing OTU clusters without impacting model performance.

We tested the ability of OptiFit to integrate new data into existing
OTUs for the purpose of machine learning classification of disease based
on microbiome abundance. A potential problem to using OptiFit for
machine learning prediction is that any sequences in the new data that
do not map to the existing OTU clusters will be discarded resulting in a
possible loss of information. However, we demonstrated that OptiFit can
be used to fit new sequence data into existing OTU clusters and perform
equally in predicting SRN compared to clustering all of the sequence
data together. The ability to integrate new data into existing OTUs
enables the deployment of a single machine learning model based on
microbiome composition that new data can be classified with. These
results are based on a single data set and disease. Further analysis is
needed to determine the data size necessary to build a robust model
capable of classifying diverse data sets. A robust machine learning
model could be implemented as part of a non-invasive and low-cost aid in
diagnosing SRN.

@@

\textbf{\emph{Data Set.}} Raw 16S rRNA amplicon sequence data isolated
from human stool samples was downloaded from NCBI Sequence Read Archive
(accession no. SRP062005) (\protect\hyperlink{ref-baxter2016}{1}).;
(\protect\hyperlink{ref-edgar2011}{6}) This data set contains stool
samples from a total of N subjects, however after preprocessing to
screen for sequence quality and subsample to 10,000 reads per sample,
490 samples remained. For this analysis, samples from subjects
identified in the metadata as normal, high risk normal, or adenoma were
categorized as ``normal'' while samples from subjects identified as
advanced adenoma or carcinoma were categorized as ``screen relevant
neoplasia'' (SRN). The resulting data set consisted of 261 normal
samples and 229 SRN samples.

\textbf{\emph{Data Processing.}} The full dataset was pre-processed with
mothur (v1.45) (\protect\hyperlink{ref-schloss2009}{7}) using the SILVA
reference database (v132) (\protect\hyperlink{ref-quast2013}{8}) to join
forward and reverse reads, merge duplicate reads, align to the
reference, pre-cluster, remove chimeras with UCHIME
(\protect\hyperlink{ref-edgar2011}{6}), assign taxonomy, and remove
non-bacterial reads following the Schloss Lab MiSeq standard operating
procedure described on the mothur website
(\url{https://mothur.org/wiki/miseq_sop/}). 100 splits of the 490
samples were generated where 80\% of the samples (392 samples) were
randomly assigned to the training set and the remaining 20\% (98
samples) were assigned to the test set. Using 100 splits of the data
accounts for the variation that may be observed depending on the samples
that are in the training or test sets. Each samples was in the training
set an average of 80 times (SD=4.1) and the test set an average of 20
times (SD=4.1).

The data was processed through two workflows. First, the standard
workflow using the OptiClust algorithm
(\protect\hyperlink{ref-westcott2017}{5}). In this pathway, all of the
data was clustered together with OptiClust to generate OTUs and the
resulting abundance tables were split into the training and testing
sets. In the second workflow, the pre-processed data was split into the
training and testing sets. The training set was clustered into OTUs,
then the test set was fit to the OTUs of the training set using the
OptiFit algorithm (\protect\hyperlink{ref-sovacool2022}{4}). The OptiFit
algorithm was run with method open so that any sequences that didn't map
to the existing OTU clusters would form new OTUs. Any OTUs that were not
in the training set were removed prior to machine learning. For both
pathways, the shared files were sub-sampled to 10,000 reads per sample.

\textbf{\emph{Machine Learning.}} Machine learning using Random Forest
was conducted with the R package mikrompl (v XXXX)
(\protect\hyperlink{ref-topuxe7uoglu2021}{9}) to predict the diagnosis
(SRN or normal) for the samples in the test set for each data split. The
training set was preprocessed to normalize values (scale/center),
collapse correlated features, and remove features with zero-variance.
The preprocessing from the training set was then applied to the test
set. P values comparing model performance were calculated as previously
described \{\}. The averaged ROC curves were plotted by taking the
average and standard deviation of the sensitivity at each specificity
value.

\textbf{\emph{Code Availability.}} The analysis workflow was implemented
in Snakemake (\protect\hyperlink{ref-koster2012}{10}) . Scripts for
analysis were written in R (\protect\hyperlink{ref-R2020}{11}) and GNU
bash (\protect\hyperlink{ref-GNUbash}{12}). The software used includes
mothur v1.47.0 (\protect\hyperlink{ref-schloss2009}{7}), RStudio
(\protect\hyperlink{ref-RStudio2019}{13}), the Tidyverse metapackage
(\protect\hyperlink{ref-wickham2019}{14}), R Markdown
(\protect\hyperlink{ref-xie_r_2018}{15}), the SRA toolkit
(\protect\hyperlink{ref-noauthor_sra-tools_nodate}{16}), and conda
(\protect\hyperlink{ref-noauthor_anaconda_2016}{17}). The complete
workflow and supporting files required to reproduce this study are
available at:
\url{https://github.com/SchlossLab/Armour_OptiFitGLNE_XXXX_2021}

@@

\leavevmode\vadjust pre{\hypertarget{ref-koster2012}{}}%
\CSLLeftMargin{10. }
\CSLRightInline{\textbf{Koster J}, \textbf{Rahmann S}. 2012.
Snakemake--a scalable bioinformatics workflow engine. Bioinformatics
\textbf{28}:2520--2522.
doi:\href{https://doi.org/10.1093/bioinformatics/bts480}{10.1093/bioinformatics/bts480}.}

\leavevmode\vadjust pre{\hypertarget{ref-R2020}{}}%
\CSLLeftMargin{11. }
\CSLRightInline{\textbf{R Core Team}. 2020.
\href{https://www.R-project.org/}{R: A language and environment for
statistical computing}. R Foundation for Statistical Computing, Vienna,
Austria.}

\leavevmode\vadjust pre{\hypertarget{ref-GNUbash}{}}%
\CSLLeftMargin{12. }
\CSLRightInline{\textbf{GNU Project}.
\href{https://www.gnu.org/software/bash/\%20manual/bash.html/}{Bash
reference manual}.}

\leavevmode\vadjust pre{\hypertarget{ref-RStudio2019}{}}%
\CSLLeftMargin{13. }
\CSLRightInline{\textbf{RStudio Team}. 2019.
\href{http://www.rstudio.com/}{RStudio: Integrated development
environment for r}. RStudio, Inc., Boston, MA.}

\leavevmode\vadjust pre{\hypertarget{ref-wickham2019}{}}%
\CSLLeftMargin{14. }
\CSLRightInline{\textbf{Wickham H}, \textbf{Averick M}, \textbf{Bryan
J}, \textbf{Chang W}, \textbf{McGowan LD}, \textbf{François R},
\textbf{Grolemund G}, \textbf{Hayes A}, \textbf{Henry L}, \textbf{Hester
J}, \textbf{Kuhn M}, \textbf{Pedersen TL}, \textbf{Miller E},
\textbf{Bache SM}, \textbf{Müller K}, \textbf{Ooms J}, \textbf{Robinson
D}, \textbf{Seidel DP}, \textbf{Spinu V}, \textbf{Takahashi K},
\textbf{Vaughan D}, \textbf{Wilke C}, \textbf{Woo K}, \textbf{Yutani H}.
2019. Welcome to the Tidyverse. Journal of Open Source Software
\textbf{4}:1686.
doi:\href{https://doi.org/10.21105/joss.01686}{10.21105/joss.01686}.}

\leavevmode\vadjust pre{\hypertarget{ref-xie_r_2018}{}}%
\CSLLeftMargin{15. }
\CSLRightInline{\textbf{Xie Y}, \textbf{Allaire JJ}, \textbf{Grolemund
G}. 2018. R {Markdown}: {The Definitive Guide}. {Taylor \& Francis, CRC
Press}.}

\leavevmode\vadjust pre{\hypertarget{ref-noauthor_sra-tools_nodate}{}}%
\CSLLeftMargin{16. }
\CSLRightInline{{SRA}-{Tools} - {NCBI}.
http://ncbi.github.io/sra-tools/.}

\leavevmode\vadjust pre{\hypertarget{ref-noauthor_anaconda_2016}{}}%
\CSLLeftMargin{17. }
\CSLRightInline{2016. Anaconda {Software Distribution}. Anaconda
Documentation. Anaconda Inc.}

>>>>

@@

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

Machine learning classification of disease based on the gut microbiome
often relies on clustering 16S rRNA gene sequences into operational
taxonomic units (OTUs) to quantify microbial composition. The standard
approach to clustering sequences into OTUs leverages the similarity of
the sequences to each other rather than to a reference database. The
abundance of each OTU is used to train a classification model. However,
OTU assignments depend on the sequences in the data set and therefore
can change if new data are added. This lack of stability complicates
classification because in order to use the model to classify additional
samples, the OTUs have to be reclustered to include the new sequences
and the model retrained with the new OTU clusters. A new reference-based
clustering algorithm, called OptiFit, addresses this issue by fitting
new sequences into existing OTUs. While OptiFit is proven to produce
high quality OTU clusters, it is unclear whether this method for fitting
new sequence data into existing OTUs will impact the performance of
classification models. We used OptiFit to cluster additional data into
existing OTU clusters and quantified model performance in classifying a
data set containing samples from patients with and without colonic
screen relevant neoplasias (SRN). We compared the performance of this
model to the standard procedure of clustering all the data together. We
found that both approaches performed equally well in classifying SRNs.
Moving forward, when OTUs are used in classification problems, OptiFit
can be used to avoid the need to retrain models using reclustered
sequences when classifying new samples.

\hypertarget{importance}{%
\subsection{Importance}\label{importance}}

There is great potential for using microbiome data to non-invasively
diagnose people. One of the challenges with using classification models
generated using the relative abundance of operational taxonomic unit
(OTU) data is that 16S rRNA gene sequences are assigned to OTUs based on
their similarity to other sequences in the dataset. If data are
generated from new patients seeking a diagnosis, then it would be
necessary to reassign sequences to OTUs and retrain the classification
model. Yet there is a desire to have a single, validated model that can
be deployed. To overcome this obstacle, we applied the OptiFit
clustering algorithm which fits new sequence data to existing OTUs
allowing the reuse of a consistent model. A random forest machine
learning model deployed using OptiFit performed as well as the
traditional reassignment and retraining approach. This result indicates
there is potential for developing machine learning classification models
based on OTU relative abundance data.

@@

Gut community composition is useful as a resource for machine learning
classification of diseases, including colorectal cancer
(\protect\hyperlink{ref-baxter2016}{1},
\protect\hyperlink{ref-zackular2014}{2}). Amplicon sequencing of the 16S
rRNA gene is a reliable tool for assessing the taxonomic composition of
microbial communities, which is the input to these models. Analysis of
16S rRNA gene sequence data generally relies on clustering of sequences
based on similarity into operational taxonomic units (OTUs). The process
of OTU clustering can either be reference-based or \emph{de novo}. The
quality of OTUs generated with reference-based clustering is generally
poor compared to those generated with \emph{de novo} clustering
(\protect\hyperlink{ref-westcott2015}{3}). While \emph{de novo}
clustering produces high-quality OTU clusters where sequences are
accurately grouped based on similarity thresholds, the resulting OTU
clusters depend on the data in the data set and the addition of new data
could change the overall OTU clusters. The unstable nature of OTU
clustering complicates deployment of machine learning models since
integration of additional data requires reclustering all the data and
retraining of the model. The ability to integrate new data into a
validated model without reclustering and retraining could allow for
deployment of a single model that new data can be continually added to.
Recently Sovacool \emph{et al} introduced OptiFit: a method for fitting
new sequence data into existing OTUs
(\protect\hyperlink{ref-sovacool2022}{4}). While OptiFit is proven to
effectively fit new sequence data to existing OTU clusters, it is
unknown if the use of OptiFit will have an impact on classification.
Here we tested the ability of OptiFit to cluster new sequence data into
existing OTU clusters for the purpose of classification of disease based
on gut microbiome composition.

We compared two approaches, one using all of the data to generate OTU
clusters and the other generating OTU clusters with a portion of the
data and then fitting the remaining sequence data to the existing OTUs
using OptiFit. In the first approach, all of the 16S rRNA sequence data
was \emph{de novo} clustered into OTUs with the OptiClust algorithm in
mothur (\protect\hyperlink{ref-westcott2017}{5}). The resulting
abundance data was then split into training and testing sets, where the
training set was used to tune hyperparameters and ultimately train the
model. The testing set was then classified with the model and the
performance of the model was quantified (Figure 1A). However, with this
methodology we would have to regenerate the OTU clusters and retrain the
model if we wanted to classify additional samples. The OptiFit algorithm
(\protect\hyperlink{ref-sovacool2022}{4}) addresses this problem by
enabling new sequences to be clustered into existing OTUs. The OptiFit
workflow is similar to the OptiClust workflow where the data was
clustered into OTUs and used to tune hyperparameters and ultimately
train the model. Then, we used OptiFit to fit sequence data of samples
not part of the original data set into the existing OTUs and used the
same model to classify the samples (Figure 1B). To test how the model
performance compares between these two methodologies, we used a publicly
available data set of 16S rRNA gene sequences from stool samples of
healthy subjects as well as subjects with SRN consisting of advanced
adenoma and carcinoma (\protect\hyperlink{ref-baxter2016}{1}). The data
set was randomly split into an 80\% train set and 20\% test set. For the
standard OptiClust workflow, the training and test sets were \emph{de
novo} clustered together into OTUs then the resulting abundance table
was split into the training and testing set. For the OptiFit workflow,
the train set was clustered \emph{de novo} into OTUs and the remaining
test set was fit to the OTU clusters using the OptiFit algorithm. For
both workflows, the abundance table of the train set was used to tune
hyperparameters and train a random forest model to classify SRN. The
test set was classified as either control or SRN using the trained
models.To account for variation depending on the split of the data, the
data set was randomly split 100 times and the process repeated for each
of the 100 data splits. By comparing the model performance of
classifying the samples in the test data set between the OptiFit and
OptiClust algorithms, we quantified the impact of using OptiFit on model
classification performance.

We first examined the quality of the resulting OTU clusters from the two
algorithms using the Matthews correlation coefficient (MCC). The MCC
score was quantified by examining all pairs of sequences and assessing
whether they belonged together in an OTU based on their similarity
(\protect\hyperlink{ref-westcott2017}{5}). MCC scores range between
negative one and one. A score of negative one means none of the
sequences in an OTU are within the similarity threshold and any
sequences within the similarity threshold are not in an OTU together. An
MCC score of zero essentially means the sequences are randomly
clustered. An MCC score of 1 means all sequences in an OTU are within
the similarity threshold and all sequence pairs within the similarity
threshold are in the same OTU. To ensure that OptiFit is appropriately
integrating new sequence data into the existing OTUs, we expected the
MCC scores produced by the OptiClust and OptiFit workflows to be
similar. Since the data was only clustered once in the OptiClust
workflow there was only one MCC score while the OptiFit workflow
produced an MCC score for the OTU clusters from each data split. Overall
the MCC scores were similar between OptiClust (MCC = 0.884) and OptiFit
(average MCC = 0.879). This indicated that OptiFit performed as well as
OptiClust when integrating new sequences into the existing OTUs.

After verifying that the quality of the OTUs was consistent between
OptiClust and OptiFit, we examined the model performance for classifying
samples in the held out test data set. To quantify model performance we
used the OTU relative abundances from the training data from the
OptiClust and OptiFit workflows to train a model to predict SRNs. Using
the predicted and actual diagnosis classification, we calculated the
area under the receiver operating characteristic curve (AUROC) for each
data split to quantify model performance. During cross-validation (CV)
training, the model performance was equivalent between the two
algorithms (p-value = 0.13, OptiClust mean CV AUROC = 0.694, OptiFit
mean CV AUROC = 0.697, Figure 2A). The trained model was then deployed
to classify the samples of the test data as control or SRN. The
performance on the test data was equivalent between the two algorithms
(p-value = 0.63, OptiClust mean test AUROC = 0.709, OptiFit mean test
AUROC = 0.712, Figure 2B,C) indicating that new data could be fit to
existing OTU clusters without impacting model performance.

We tested the ability of OptiFit to integrate new data into existing
OTUs for the purpose of machine learning classification using OTU
relative abundance. A potential problem with using OptiFit is that any
sequences in the new data that do not map to the existing OTU clusters
will be discarded resulting in a possible loss of information. However,
we demonstrated that OptiFit can be used to fit new sequence data into
existing OTU clusters and perform equally well in predicting SRN
compared to clustering all of the sequence data together. The ability to
integrate data from new samples into existing OTUs enables the
deployment of a single machine learning model. These results are based
on a single data set and disease. Further analysis is needed to
determine the number of samples that are necessary to build a robust
model capable of classifying diverse samples. A robust machine learning
model could be implemented as part of a non-invasive and low-cost aid in
diagnosing SRN and other diseases.

@@

\textbf{\emph{Data Set.}} Raw 16S rRNA gene sequence data isolated from
human stool samples was downloaded from NCBI Sequence Read Archive
(accession no. SRP062005) (\protect\hyperlink{ref-baxter2016}{1},
\protect\hyperlink{ref-edgar2011}{6}). This data set contains stool
samples from a total of 490 subjects. For this analysis, samples from
subjects identified in the metadata as normal, high risk normal, or
adenoma were categorized as ``normal'' while samples from subjects
identified as advanced adenoma or carcinoma were categorized as ``screen
relevant neoplasia'' (SRN). The resulting data set consisted of 261
normal samples and 229 SRN samples.

\textbf{\emph{Data Processing.}} The full dataset was preprocessed with
mothur (v1.47) (\protect\hyperlink{ref-schloss2009}{7}) to join forward
and reverse reads, merge duplicate reads, align to the SILVA reference
database (v132) (\protect\hyperlink{ref-quast2013}{8}), precluster,
remove chimeras with UCHIME (\protect\hyperlink{ref-edgar2011}{6}),
assign taxonomy, and remove non-bacterial reads following the Schloss
Lab MiSeq standard operating procedure described on the mothur website
(\url{https://mothur.org/wiki/miseq_sop/}). 100 splits of the 490
samples were generated where 80\% of the samples (392 samples) were
randomly assigned to the training set and the remaining 20\% (98
samples) were assigned to the test set. Using 100 splits of the data
accounts for the variation that may be observed depending on the samples
that are in the training or test sets. Each sample was in the training
set an average of 80 times (SD=4.1) and the test set an average of 20
times (SD=4.1).

The data was processed through two workflows. First, the standard
workflow using the OptiClust algorithm
(\protect\hyperlink{ref-westcott2017}{5}). In this pathway, all of the
data was clustered together with OptiClust to generate OTUs and the
resulting abundance tables were split into the training and testing
sets. In the second workflow, the preprocessed data was split into the
training and testing sets. The training set was clustered into OTUs,
then the test set was fit to the OTUs of the training set using the
OptiFit algorithm (\protect\hyperlink{ref-sovacool2022}{4}). The OptiFit
algorithm was run with method open so that any sequences that did not
map to the existing OTU clusters would form new OTUs. For both pathways,
the shared files were sub-sampled to 10,000 reads per sample.

\textbf{\emph{Machine Learning.}} Machine learning using Random Forest
was conducted with the R package mikrompl (v 1.2.0)
(\protect\hyperlink{ref-topuxe7uoglu2021}{9}) to predict the diagnosis
(SRN or normal) for the samples in the test set for each data split. The
training set was preprocessed to normalize OTU counts (scale/center),
collapse correlated OTUs, and remove OTUs with zero-variance. The
preprocessing from the training set was then applied to the test set.
Any OTUs in the test set that were not in the training set were removed.
P values comparing model performance were calculated as previously
described (\protect\hyperlink{ref-topuxe7uoglu2020}{10}). The averaged
ROC curves were plotted by taking the average and standard deviation of
the sensitivity at each specificity value.

\textbf{\emph{Code Availability.}} The analysis workflow was implemented
in Snakemake (\protect\hyperlink{ref-koster2012}{11}) . Scripts for
analysis were written in R (\protect\hyperlink{ref-R2020}{12}) and GNU
bash (\protect\hyperlink{ref-GNUbash}{13}). The software used includes
mothur v1.47.0 (\protect\hyperlink{ref-schloss2009}{7}), RStudio
(\protect\hyperlink{ref-RStudio2019}{14}), the Tidyverse metapackage
(\protect\hyperlink{ref-wickham2019}{15}), R Markdown
(\protect\hyperlink{ref-xie_r_2018}{16}), the SRA toolkit
(\protect\hyperlink{ref-noauthor_sra-tools_nodate}{17}), and conda
(\protect\hyperlink{ref-noauthor_anaconda_2016}{18}). The complete
workflow and supporting files required to reproduce this study are
available at:
\url{https://github.com/SchlossLab/Armour_OptiFitGLNE_XXXX_2021}

@@

\leavevmode\vadjust pre{\hypertarget{ref-topuxe7uoglu2020}{}}%
\CSLLeftMargin{10. }
\CSLRightInline{\textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
\textbf{Ruffin MT}, \textbf{Wiens J}, \textbf{Schloss PD}. 2020. A
framework for effective application of machine learning to
microbiome-based classification problems. mBio \textbf{11}:e00434--20.
doi:\href{https://doi.org/10.1128/mBio.00434-20}{10.1128/mBio.00434-20}.}

\leavevmode\vadjust pre{\hypertarget{ref-koster2012}{}}%
\CSLLeftMargin{11. }
\CSLRightInline{\textbf{Koster J}, \textbf{Rahmann S}. 2012.
Snakemake--a scalable bioinformatics workflow engine. Bioinformatics
\textbf{28}:2520--2522.
doi:\href{https://doi.org/10.1093/bioinformatics/bts480}{10.1093/bioinformatics/bts480}.}

\leavevmode\vadjust pre{\hypertarget{ref-R2020}{}}%
\CSLLeftMargin{12. }
\CSLRightInline{\textbf{R Core Team}. 2020.
\href{https://www.R-project.org/}{R: A language and environment for
statistical computing}. R Foundation for Statistical Computing, Vienna,
Austria.}

\leavevmode\vadjust pre{\hypertarget{ref-GNUbash}{}}%
\CSLLeftMargin{13. }
\CSLRightInline{\textbf{GNU Project}.
\href{https://www.gnu.org/software/bash/\%20manual/bash.html/}{Bash
reference manual}.}

\leavevmode\vadjust pre{\hypertarget{ref-RStudio2019}{}}%
\CSLLeftMargin{14. }
\CSLRightInline{\textbf{RStudio Team}. 2019.
\href{http://www.rstudio.com/}{RStudio: Integrated development
environment for r}. RStudio, Inc., Boston, MA.}

\leavevmode\vadjust pre{\hypertarget{ref-wickham2019}{}}%
\CSLLeftMargin{15. }
\CSLRightInline{\textbf{Wickham H}, \textbf{Averick M}, \textbf{Bryan
J}, \textbf{Chang W}, \textbf{McGowan LD}, \textbf{François R},
\textbf{Grolemund G}, \textbf{Hayes A}, \textbf{Henry L}, \textbf{Hester
J}, \textbf{Kuhn M}, \textbf{Pedersen TL}, \textbf{Miller E},
\textbf{Bache SM}, \textbf{Müller K}, \textbf{Ooms J}, \textbf{Robinson
D}, \textbf{Seidel DP}, \textbf{Spinu V}, \textbf{Takahashi K},
\textbf{Vaughan D}, \textbf{Wilke C}, \textbf{Woo K}, \textbf{Yutani H}.
2019. Welcome to the Tidyverse. Journal of Open Source Software
\textbf{4}:1686.
doi:\href{https://doi.org/10.21105/joss.01686}{10.21105/joss.01686}.}

\leavevmode\vadjust pre{\hypertarget{ref-xie_r_2018}{}}%
\CSLLeftMargin{16. }
\CSLRightInline{\textbf{Xie Y}, \textbf{Allaire JJ}, \textbf{Grolemund
G}. 2018. R {Markdown}: {The Definitive Guide}. {Taylor \& Francis, CRC
Press}.}

\leavevmode\vadjust pre{\hypertarget{ref-noauthor_sra-tools_nodate}{}}%
\CSLLeftMargin{17. }
\CSLRightInline{{SRA}-{Tools} - {NCBI}.
http://ncbi.github.io/sra-tools/.}

\leavevmode\vadjust pre{\hypertarget{ref-noauthor_anaconda_2016}{}}%
\CSLLeftMargin{18. }
\CSLRightInline{2016. Anaconda {Software Distribution}. Anaconda
Documentation. Anaconda Inc.}
